---
title: "Reproducible Research Class Project Week 2"
author: "Maria Telpoukhovskaia"
date: "November 18, 2016"

output: html_document
keep_md: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Introduction to the project

In this project, we examine the data set with steps taken by a person wearing an activity monitoring device in 5 minute increments in October and November in 2012.

## 1. Loading data

The activity data analyzed is first loaded.

```{r}
activity <- read.csv("activity.csv", stringsAsFactors = FALSE)
head(activity)
```

Date is converted from character to Date class.
```{r}
activity$date <- as.Date(activity$date)
```

##2. What is the mean total number of steps taken per day?

First, a data frame with daily total is created with aggregate function.

```{r}
total.steps.per.day <- aggregate(steps~date, activity, sum)
```

Second, we examine the histogram of total steps daken per day.

```{r}
library(ggplot2)
g <- ggplot(total.steps.per.day, aes(x=steps))
g <- g + geom_histogram(binwidth = 250, fill = 'darkgreen', colour = "white")
g <- g + ggtitle("Histogram of Total Steps Taken Per Day")
g <- g + theme(plot.title = element_text (size=20, face="bold"))
g <- g + theme(axis.text.x = element_text(size = 20))
g <- g + labs(x = "# of Steps Taken")
g <- g + theme(axis.title.x = element_text(size = 20))
g <- g + theme(axis.title.y = element_text(size = 20))
g <- g + labs(y = "Count")
g <- g + theme(axis.text.y = element_text(size = 20))
g
```

Third, we calculate the mean and median for this dataset.

```{r}
mean(total.steps.per.day$steps)
median(total.steps.per.day$steps)
```


##3. What is the average daily activity pattern?

First, we make a data frame with average steps per interval.

```{r}
average.steps.per.interval <- aggregate(steps~interval, activity, mean)
```


Second, we plot a time series plot of the 5-minute interval and the average number of steps.

```{r}
plot(average.steps.per.interval$interval, average.steps.per.interval$steps, 
     type="l",
     col = "darkblue",
     xlab = "5 Minute Interval",
     ylab = "Average Steps Per Interval")
title("Average Steps Per Interval")
```

Last, we calculate which 5-minute interval has the maximum number of steps.

```{r, message=FALSE}
library(dplyr)
summarise_each(average.steps.per.interval, funs(max(., na.rm=TRUE)))
```

##4. Imputing missing values

There are a number of missing values in the steps column.  First, we calculate that number.

```{r}
missing.steps <- sum(is.na(activity$steps))
missing.steps
```

We find that there are `r missing.steps` number of missing values.

Second, we'll fill these missing values using the mean for that interval using the plyr package.

```{r, message=FALSE}
library(plyr)
impute.mean <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))
activity2 <- ddply(activity, ~ interval , transform, steps = impute.mean(steps))
activity2 <- activity2[order(activity2$date), ]
```

We investigate this data set with a histogram for the total number of steps taken each day.

```{r}
total.steps.per.day2 <- aggregate(steps~date, activity2, sum)
g <- ggplot(total.steps.per.day2, aes(x=steps))
g <- g + geom_histogram(binwidth = 250, fill = 'darkgreen', colour = "white")
g <- g + ggtitle("Histogram of Total Steps Taken Per Day \n (missing values filled in)")
g <- g + theme(plot.title = element_text (size=20, face="bold"))
g <- g + theme(axis.text.x = element_text(size = 20))
g <- g + labs(x = "# of Steps Taken")
g <- g + theme(axis.title.x = element_text(size = 20))
g <- g + theme(axis.title.y = element_text(size = 20))
g <- g + labs(y = "Count")
g <- g + theme(axis.text.y = element_text(size = 20))
g
```

The mean and median for this data set is computed.

```{r}
mean.s <- mean(total.steps.per.day$steps)
median.s <- median(total.steps.per.day$steps)
mean.s2 <- mean(total.steps.per.day2$steps)
median.s2 <- median(total.steps.per.day2$steps)
mean.difference <- mean.s - mean.s2
median.difference <- median.s - median.s2
```

The mean and the median for the new data set are `r mean.s2` and `r median.s2`, respectively.  The first mean and median are `r mean.s` and `r median.s`, respectively.  By filling in the missing values, the mean changes by `r mean.difference`, and the median by `r median.difference`.

##5. Are there differences in activity patterns between weekdays and weekends?

First, we create a new data set with a new column that indicates whether a day is a weekday or a weekend using the function mutate.

```{r}
weekend <- c("Saturday", "Sunday")

activity3 <- mutate(activity2,
                    day = factor(1 * (weekdays(activity2$date) %in% weekend),
                    labels = c("weekday", "weekend")))
```


We can now examine average activity for weekend and weekday days, and see that weekday has a spike in activity.

```{r}
average.steps.per.interval2 <- aggregate(steps~day+interval, activity3, mean)

g <- ggplot(average.steps.per.interval2, aes(x = interval, y = steps)) +
     geom_line(lwd = 2, col = "darkred")
g <- g + ggtitle("Weekday vs. Weekend Activity")
g <- g + theme(plot.title = element_text (size=20, face="bold"))
g <- g + theme(axis.text.x = element_text(size = 20, angle = 45))
g <- g + labs(x = "Inverval")
g <- g + theme(axis.title.x = element_text(size = 20))
g <- g + theme(axis.title.y = element_text(size = 20))
g <- g + labs(y = "# of Steps Taken")
g <- g + theme(axis.text.y = element_text(size = 20))
g <- g + facet_grid(. ~ day)
g
```
